---
title: "Politics Are Afoot!"
author: 'Eda Kavlakoglu'
output: pdf_document
---

# The Setup 

There is *a lot* of money that is spent in politics in Presidential election years. So far, estimates have the number at about $11,000,000,000 (11 billion USD). For context, in 2019 Twitter's annual revenue was about \$3,500,000,000 (3.5 billion USD). 

# The work 

Install the package, `fec16`. 

```{r}
install.packages('fec16')
```

This package is a compendium  of spending and results from the 2016 election cycle. In this dataset are 9 different datasets that cover: 

- `candidates`: candidate attributes, like their name, a unique id of the candidate, the election year under consideration, the office they're running for, etc. 
- `results_house`: race attributes, like the name of the candidates running in the election, a unique id of the candidate, the number of `general_votes` garnered by each candidate, and other information. 
- `campaigns`: financial information for each house & senate campaign. This includes a unique candidate id, the total receipts (how much came in the doors), and total disbursements (the total spent by the campaign), the total contributed by party central committees, and other information. 

# Your task 

Describe the relationship between spending on a candidate's behalf and the votes they receive.

# Your work 

- We want to keep this work *relatively* constrained, which is why we're providing you with data through the `fec16` package. It is possible to gather all the information from current FEC reports, but it would require you to make a series of API calls that would pull us away from the core modeling tasks that we want you to focus on instead. 
- Throughout this assignment, limit yourself to functions that are  within the `tidyverse` family of packages: `dplyr`, `ggplot`, `patchwork`, and `magrittr` for wrangling and exploration and `base`, `stats`, `sandwich` and `lmtest` for modeling and testing. You do not *have* to use these packages; but try to limit yourself to using only these. 

```{r load packages, message=FALSE}
library(tidyverse)
library(magrittr)
library(ggplot2)
library(patchwork)
library(sandwich)
library(lmtest)
library(fec16)
theme_set(theme_minimal())
knitr::opts_chunk$set(dpi = 300)
```

```{r load data}
candidates    <- fec16::candidates
results_house <- fec16::results_house
campaigns     <- fec16::campaigns
```

## 1. What does the distribution of votes and of spending look like? 

1. (3 points) In separate histograms, show both the distribution of votes (measured in `results_house$general_percent` for now) and spending (measured in `ttl_disb`).  Use a log transform if appropriate for each visualization.  How would you describe what you see in these two plots?

```{r}
#filter out NA values
f_rh <- results_house %>% 
  drop_na(general_percent, general_votes)
  #filter(!is.na(general_votes))

#plot histogram
votes_histogram <- f_rh %>% 
  ggplot() + 
  aes(x = general_percent) + 
  geom_histogram(bins = 50) +
  scale_x_continuous(name = 'Percent of Votes Won', labels = scales::percent) +
  labs(
    title    = 'Fig. 1: Distribution of Votes by Percent', 
    subtitle = 'The shape of voting distribution is fairly trimodal with the highest peak at low or no votes', 
    y = 'count of candidates'
  ) 

# filter negative campaign spend
f_cpmgn <- campaigns %>%
  filter(ttl_disb>=0) 

#log spending data and add 1 to every campaign to enable log transform for campaigns with 0 spend
  f_cpmgn$log_ttl_disb=log(f_cpmgn$ttl_disb +1)
  f_rh$log_general_votes=log(f_rh$general_votes +1)

#plot histogram
spending_histogram <- f_cpmgn %>% 
  ggplot() + 
  aes(x = log_ttl_disb) + 
  geom_histogram(bins = 50) +
  scale_x_continuous(name = 'Log of Campaign Spend ($)', labels= scales::dollar) + 
  labs(
    title    = 'Fig. 2: Distribution of Campaign Spend (with Log Transform)', 
    subtitle = 'The shape of campaign spend distribution is mostly bimodal with an outlier at zero spend', 
    y        = 'Count of Candidates'
  ) 
votes_histogram / spending_histogram
```
Neither of these plots seem to have a normal distribution. In Figure 1, the distribution appears trimodal. The first peak occurs at the lowest percentages of votes won. Based on the distribution of Figure 2, we could potentially hypothesize that candidates with low or no campaign spending did not receive many votes; however, there could be other factors that may contribute to this peak, such as political party. The other two peaks in Figure 1 demonstrate that candidates received a range of 20 to 50 percent of votes or 50 to 80 percent of votes, which seems to reflect the "closely contested" elections and "uncontested victories" noted in the Mayhew article.

In Figure 2, we transform the campaign spend data with a log to show a more normal distribution. While you might be able to argue that Figure 2 is also trimodal, I'd err on the side of bimodal with significant outliers at zero. When we focus on the primary distribution of the data, it shows that there's two groups of candidates--one which has a higher amount of campaign spend, ranging between 7 and 13 dollars, and another one, ranging from 13 and 16 dollars.

## 2. Exploring the relationship between spending and votes. 

2. (3 points) Create a new dataframe by joining `results_house` and `campaigns` using the `inner_join` function from `dplyr`. (We use the format `package::function` -- so `dplyr::inner_join`.) 


```{r}
new_df <- inner_join(f_rh, f_cpmgn, by = 'cand_id')
```


3. (3 points) Produce a scatter plot of `general_votes` on the y-axis and `ttl_disb` on the x-axis. What do you observe about the shape of the joint distribution? 

Figure 3: Much of the campaign spend and voting behavior is clustered in the left bottom corner of the scatterplot, showing that a subset of candidates receive around 100K to 300K votes based on some campaign spending. Candidates with higher campaign spend tend to receive more votes, but this seems to hit a diminishing return around $2M. This observation also does not seem to be true for all candidates. Since there is a subset of candidates that do not seem to perform well with higher campaign spend, there must be another important relationship, which helps to explain voting behavior.

```{r}
vote_spend_scatter <- new_df %>%
  ggplot()+
  aes(x=ttl_disb, y=general_votes) + 
  scale_x_continuous(name = 'Campaign Spend ($)', labels= scales::dollar) + 
  scale_y_continuous(name = 'Count of Votes', labels = scales::comma) + 
  geom_point() +
  labs(
    title    = 'Fig. 3: Understanding the relationship between campaign spend and voting', 
    subtitle = 'Candidates with higher campaign spend receive more votes but eventually hits diminishing returns'
  ) 

vote_spend_scatter
```
Figure 4: This relationship can be more clearly seen when we take the log10 of campaign spend. Candidates with higher campaign spend tend to receive more votes, but again, this doesn't seem to be true of all candidates--i.e. some candidates with high campaign spend received little to no votes. This indicates another important relationship in the data. 

```{r}
vote_spend_scatter <- new_df %>%
  ggplot()+
  aes(x=log_ttl_disb, y=general_votes) + 
  scale_x_continuous(name = 'Log of Campaign Spend ($)', labels= scales::dollar) + 
  scale_y_continuous(name = 'Count of Votes', labels = scales::comma) + 
  geom_point() +
  labs(
    title    = 'Fig. 4: Understanding the relationship between campaign spend and voting', 
    subtitle = 'Candidates with higher campaign spend tend to receive more votes but not for all of them'
  ) 

vote_spend_scatter
```

4. (3 points) Create a new variable to indicate whether each individual is a "Democrat", "Republican" or "Other Party". 
  - Here's an example of how you might use `mutate` and `case_when` together to create a variable. 

```
starwars %>%
  select(name:mass, gender, species) %>%
  mutate(
  type = case_when(
    height > 200 | mass > 200 ~ "large",
    species == "Droid"        ~ "robot",
    TRUE                      ~ "other"
    )
  )
```  
```{r}  
df <- new_df %>%
  mutate(
    party_name = case_when(
      party == "REP" ~ "Republican", 
      party == "DEM" ~ "Democrat", 
      TRUE ~ "Other Party"
    ), 
    is_rep = case_when(
      party == "REP" ~ 1, 
      party != "REP" ~ 0
    ), 
    is_dem = case_when(
      party == "DEM" ~ 1, 
      party != "DEM" ~ 0
    ), 
    is_other = case_when(
      party_name == "Other Party" ~ 1, 
      TRUE ~ 0
    )
  )
```

Once you've produced the new variable, plot your scatter plot again, but this time adding an argument into the `aes()` function that colors the points by party membership.  What do you observe about the distribution of all three variables?

3rd party candidates struggle to win votes 

```{r}
df %>%
  ggplot() +
  aes(x=ttl_disb, y=general_votes, color = party_name) + 
  geom_point() +
  scale_x_continuous(name = 'Campaign Spend ($)', labels= scales::dollar) + 
  scale_y_continuous(name = 'Count of Votes', labels = scales::comma) + 
  labs(
    title    = 'Fig. 5: The relationship between campaign spend and voting by party', 
    subtitle = '3rd party candidates struggle to get votes even with sizeable campaign spend', 
    color = 'political party'
  ) 

```

```{r}
df %>%
  ggplot() +
  aes(x=log_ttl_disb, y=general_votes, color = party_name) + 
  geom_point() +
  scale_x_continuous(name = 'Log of Campaign Spend ($)', labels= scales::dollar) + 
  scale_y_continuous(name = 'Count of Votes', labels = scales::comma) + 
  labs(
    title    = 'Fig. 6: Understanding the relationship between campaign spend and voting by party', 
    subtitle = '3rd party candidates struggle to get votes even with sizeable campaign spend', 
    color = 'political party'
  ) 
```
# Produce a Descriptive Model 

5. (5 Points) Given your observations, produce a linear model that you think does a good job at describing the relationship between candidate spending and votes they receive. You should decide what transformation to apply to spending (if any), what transformation to apply to votes (if any) and also how to include the party affiliation.

$$
log(general\ votes)_{i} = \beta_0 + \beta_1 Republicans_{i} + \beta_2 OtherParty_{i} + \beta_3 log(CampaignSpend_{i}) + \beta_4 log(CampaignSpend_{i} \times OtherParty_{i}) + \epsilon_{i}
$$

```{r}
short_model <- lm(log_general_votes ~ is_rep + is_other + log_ttl_disb + (is_other*log_ttl_disb), data = df)
long_model<- lm(log_general_votes ~ is_rep + is_other + log_ttl_disb + (is_other*log_ttl_disb) + (is_rep*log_ttl_disb), data = df)
summary(short_model)
summary(long_model)
```

```{r}
vote_spend_scatter_lm <- df %>%
  ggplot()+
  aes(x=log_ttl_disb, y=log_general_votes, color=party_name) + 
  scale_x_continuous(name = 'Log of Campaign Spend ($)', labels= scales::dollar) + 
  scale_y_continuous(name = 'Log of Vote Count', labels = scales::comma) + 
  geom_point() +
  geom_smooth(method = "lm", formula = "y~x", se = FALSE) + 
  labs(
    title    = 'Fig. 7: Understanding the relationship between campaign spend and voting', 
    subtitle = 'Insert insight here',
    color = 'political party'
  ) 
vote_spend_scatter_lm
```

```{r}
#do I need to add another interaction term for the other party category (i.e Republican*log_ttl_disb)
anova(short_model, long_model, test = "F")
#yes, use the long-model since the p-value is less than 0.05.
```

6. (3 points) Evaluate the Large-Sample Linear Model Assumptions

After we log both the general vote data and the campaign spend data, the distribution of the data looks more normal/well behaved. While the data itself may not be IID since candidates in one election can influence the candidates in another election, we'll proceed with this analysis under the assumptions of a large-sample model are met, meaning that (1) the data is IID, (2) the data is well-behaved data and large enough to allow CLT to kick in, and (3) unique BLP exists. 

7. (3 points) Interpret the model coefficients you estimate.

- Tasks to keep in mind as you're writing about your model: 
    - At the time that you're writing and interpreting your regression coefficients you'll be *deep* in the analysis. Nobody will know more about the data than you do, at that point. *So, although it will feel tedious, be descriptive and thorough in describing your observations.* 
    - It can be hard to strike the balance between: on the one hand,  writing enough of the technical underpinnings to know that your model meets the assumptions that it must; and, on the other hand, writing little enough about the model assumptions that the implications of the model can still be clear. We're starting this practice now, so that by the end of Lab 2 you will have had several chances to strike this balance.

- When campaign spend is equal to 0, a Democratic candidate can expect to win 36.3K votes (i.e. e^10.5). 
- Holding all other variables constant, a Republican candidate will get 83.5% less votes than a Democratic candidate.
- Holding all other variables constant, a candidate from an "Other Party" will get 160% less votes than a Democratic candidate.
-For models with Democratic candidate, campaign spend would yield 10% more votes than if the candidate did not have campaign spend, holding all other variables constant. In models with a Republican candidate, campaign spend would increase votes won by 6.4% while campaign spend will actually decrease votes won by 7.3% for models with an other party candidate. 
-All of the variables in this model are considered statistically significant, meaning that political party and campaign spend contribute to the amount of votes won in an election. 